Structuring College Access: The Market Segment Model and College Board Geomarkets

### 2025-04-16 text/back_end_KS.qmd
dallas maps rq1

These population shifts have contributed to changes in the racial/ethnic composition within and across Geomarkets over time. @fig-chicago-rq1-race illustrates the racial and ethnic composition of each Geomarkets across the three census periods. In 1980, all Geomarkets —except for the City of Chicago - were predominantly White, with proportions ranging from 83.1% in the South and Southwest Suburbs to approximately 93.3% in North Shore. In contrast, the City of Chicago had a substantially more diverse population, with White residents comprising only 45.5% of its total population. By 2020, several Geomarkets that were previously predominantly White shifted to a more balanced racial/ethnic distribution, including Chain of Lakes (58.7% White, 7.6% Black, 23.5% Hispanic; 7.4% Asian Pacific Islander), Northwest Suburbs (61.5% White, 3% Black, 18.2% Hispanic; 14.8% Asian Pacific Islander), Evanston and Skokie (63.7% White, 6.7% Black, 10% Hispanic; 16% Asian Pacific Islander), Western Suburbs (55.6% White, 9.3% Black, 23.4% Hispanic; 9.2% Asian Pacific Islander), and South and Southwest Suburbs (53.4% White, 23.8% Black, 17.9% Hispanic; 2.8% Asian Pacific Islander). For many Geomarkets, this balance in racial/ethnic composition was the result of large increases in the Latinx population from 1980 to 2020, which ranged from a 7.4 percentage point increase for the Evanston and Skokie Geomarket to a 19.2 percentage point increase for Western Suburbs. Some Geomarkets also experienced a substantial increase in Black populations (South and Southwest Suburbs) and in Asian populations (Northwest Suburbs and Evanston and Skokie).

@fig-chicago-rq1-race also indicates that, although the North Shore and City of Chicago Geomarkets underwent some changes in racial and ethnic composition, their overall demographic patterns remained relatively the same across the three census periods. For instance, North Shore experienced some increases in Hispanic (2.2.% in 1980 to 7.9% in 2020), Asian Pacific Islander (6.6% in 2000 to 9.9% in 2020), and Black (1.8.% in 1980 to 2.6% in 2020) populations yet remained a predominantly White Geomarket (74% in 2020). City of Chicago also experienced relatively modest declines in its Black population (36.6% in 1980 to 32.9% in 2020), alongside increases in its Hispanic (14.8% in 1980 to 25.8% in 2020) and Asian Pacific Islander populations (4.0% in 2000 to 6.1% in 2020), further substantiating the Geomarket as the most racially/ethnic diverse in the metropolitan area.

@@

@fig-chicago-rq1-ses also shows that from 1980 to 2020, the percentage of residents living in poverty declined in the City of Chicago Geomarket while it increased in all other Geomarkets, resulting in a narrowing of poverty gaps across the Chicago metropolitan area. Similar patterns are evident in educational attainment across Geomarkets. In 1980, the City of Chicago had the lowest percentage of residents with a bachelor's degree or higher (12.5%). However, by 2020, the Geomarket experienced the largest percentage point increase in college degree attainment, reaching 37.6%. In comparison, all other Geomarkets experienced smaller gains in the share of residents with a bachelor’s degree or higher over the same period. These trends contributed to a narrowing of the educational attainment gap across Geomarkets from 1980 to 2020.

@@

>>>>

These population shifts have contributed to changes in the racial/ethnic composition within and across Geomarkets over time. @fig-chicago-rq1-race illustrates the racial and ethnic composition of each Geomarkets across the three census periods. In 1980, all Geomarkets —except for the City of Chicago - were predominantly White, with proportions ranging from 83% in the South and Southwest Suburbs to approximately 93% in North Shore. In contrast, the City of Chicago had a substantially more diverse population, with White residents comprising only 46% of its total population. By 2020, several Geomarkets that were previously predominantly White shifted to a more balanced racial/ethnic distribution, including Chain of Lakes (59% White, 8% Black, 24% Hispanic; 7% Asian Pacific Islander), Northwest Suburbs (62% White, 3% Black, 18% Hispanic; 15% Asian Pacific Islander), Evanston and Skokie (64% White, 7% Black, 10% Hispanic; 16% Asian Pacific Islander), Western Suburbs (56% White, 9% Black, 23% Hispanic; 9% Asian Pacific Islander), and South and Southwest Suburbs (53% White, 24% Black, 18% Hispanic; 3% Asian Pacific Islander). For many Geomarkets, this balance in racial/ethnic composition was the result of large increases in the Latinx population from 1980 to 2020, which ranged from a 7 percentage point increase for the Evanston and Skokie Geomarket to a 19 percentage point increase for Western Suburbs. Some Geomarkets also experienced a substantial increase in Black populations (South and Southwest Suburbs) and in Asian populations (Northwest Suburbs and Evanston and Skokie).

@fig-chicago-rq1-race also indicates that, although the North Shore and City of Chicago Geomarkets underwent some changes in racial and ethnic composition, their overall demographic patterns remained relatively the same across the three census periods. For instance, North Shore experienced some increases in Hispanic (2% in 1980 to 8% in 2020), Asian Pacific Islander (7% in 2000 to 10% in 2020), and Black (2% in 1980 to 3% in 2020) populations yet remained a predominantly White Geomarket (74% in 2020). City of Chicago also experienced relatively modest declines in its Black population (37% in 1980 to 33% in 2020), alongside increases in its Hispanic (15% in 1980 to 26% in 2020) and Asian Pacific Islander populations (4% in 2000 to 6% in 2020), further substantiating the Geomarket as the most racially/ethnic diverse in the metropolitan area.

@@

@fig-chicago-rq1-ses also shows that from 1980 to 2020, the percentage of residents living in poverty declined in the City of Chicago Geomarket while it increased in all other Geomarkets, resulting in a narrowing of poverty gaps across the Chicago metropolitan area. Similar patterns are evident in educational attainment across Geomarkets. In 1980, the City of Chicago had the lowest percentage of residents with a bachelor's degree or higher (13%). However, by 2020, the Geomarket experienced the largest percentage point increase in college degree attainment, reaching 38%. In comparison, all other Geomarkets experienced smaller gains in the share of residents with a bachelor’s degree or higher over the same period. These trends contributed to a narrowing of the educational attainment gap across Geomarkets from 1980 to 2020.

@@

For Dallas-Fort Worth, Figure X illustrates the racial and ethnic composition of each Geomarket across the three census periods. In 1980, all Geomarkets were predominantly White, with proportions ranging from 62% in City of Dallas to nearly 92% in West of Dallas/Ft. Worth Metroplex. However, by 2000, Geomarkets located in the central part of the metropolitan area shifted to a more balanced racial/ethnic distribution, including Dallas Co (58% White, 16% Black, 19% Hispanic; 6% Asian Pacific Islander), Irving, Arlington, and Grand Prairie (54% White, 12% Black, 26% Hispanic; 16% Asian Pacific Islander), and City of Fort Worth (57% White, 16% Black, 22% Hispanic; 3% Asian Pacific Islander). For the City of Dallas Geomarket,
